Determinism, Complexity, and Predictability in Computer Performance

### 2014-02-10 pecomplexity.tex
add figures to introduction

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:cache}, and chaos places fundamental limits on
predictability.
%
% \begin{figure}[htbp]
%    \centering
%    \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
%    % where an .eps filename suffix will be assumed under latex,
%    % and a .pdf suffix will be assumed for pdflatex
%    \caption{A small snippet of the L2 cache miss rate of {\tt
%        col\_major}, a three-line C program that repeatedly initializes
%      a matrix in column-major order, running on an Intel Core
%      Duo\textsuperscript{\textregistered}-based machine.  Even this
%      simple program exhibits chaotic performance dynamics.}
%    \label{fig:cache}
%  \end{figure}

>>>>

Computers are among the most complex engineered artifacts in current
use.  Modern microprocessor chips contain multiple processing units
and multi-layer memories, for instance, and they use complicated
hardware/software strategies to move data and threads of computation
across those resources.  These features---along with all the others
that go into the design of these chips---make the patterns of their
processor loads and memory accesses highly complex and hard to
predict.  Accurate forecasts of these quantities, if one could
construct them, could be used to improve computer design.  If one
could predict that a particular computational thread would be bogged
down for the next 0.6 seconds waiting for data from main memory, for
instance, one could save power by putting that thread on hold for that
time period (e.g., by migrating it to a processing unit whose clock
speed is scaled back).  Computer performance traces are, however, very
complex.  Even a simple ``microkernel,'' like a three-line loop that
repeatedly initializes a matrix in column-major order, can produce
{\sl chaotic} performance traces \cite{mytkowicz09}, as shown in
Figure~\ref{fig:col-ts}, and chaos places fundamental limits on
predictability.
%
% \begin{figure}[htbp]
%    \centering
%    \includegraphics[width=0.5\textwidth]{figs/colCacheShort}
%    % where an .eps filename suffix will be assumed under latex,
%    % and a .pdf suffix will be assumed for pdflatex
%    \caption{A small snippet of the L2 cache miss rate of {\tt
%        col\_major}, a three-line C program that repeatedly initializes
%      a matrix in column-major order, running on an Intel Core
%      Duo\textsuperscript{\textregistered}-based machine.  Even this
%      simple program exhibits chaotic performance dynamics.}
%    \label{fig:cache}
%  \end{figure}

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/colshortts}
    \caption{A short segment of the instructions executed per CPU clock cycle
    (IPC) during the execution of \col. Each point is the IPC during a 100,000
    cycle period. There is a great deal of periodicity in this time series.}
    \label{fig:col-ts}
  \end{subfigure}%
  \begin{subfigure}[t]{0.475\textwidth}
    \includegraphics[width=0.95\textwidth]{figs/gccfullts}
    \caption{The full IPC time series during the execution of \gcc. This time
    series has very little structure.}
    \label{fig:gcc-ts}
  \end{subfigure}
  \label{fig:sample-ts}
\end{figure}
